Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.

### 2023-11-27 article.qmd
Move code blocks to a figure and add mathematical expression

Poincare is a package to define and simulate dynamical systems.
It provides a `System` class,
where we can define `Constant`s, `Parameter`s, `Variable`s,
and create equations linking a variable's derivative with an expression:

```{python}
from poincare import Variable, System, initial

class Decay(System):
    x: Variable = initial(default=1)  # Variable with initial condition
    eq = x.derive() << -x  # equation assigning dx/dt = -x
```

It also allows to define higher-order systems
by assigning an initial condition to a `Derivative`:

```{python}
from poincare import Derivative

class Oscillator(System):
    x: Variable = initial(default=1)
    v: Derivative = x.derive(initial=0)
    eq = v.derive() << -x
```

@@

1. the variable name to which a component is assigned can be automatically saved in the component for introspection (i.e., `Oscilator.x.name == "x"`),
2. it provides a namespace such that we can easily define multiple independent models in the same script,
3. it allows IDEs to provide autocomplete and refactoring capabilities (`Oscillator.<TAB>` shows `x`, `v` and `eq`),
4. it allows to create instances which can be composed into a bigger model.

@@

```{python}
class BigModel(System):
    x: Variable = initial(default=1)
    subsystem = Decay(x=x)
    independent = Decay(x=1)  # independent variable created inside
```

>>>>

```{python}
# | echo: false
import numpy as np
import seaborn
from poincare import Derivative, Variable, System, initial
from poincare import Simulator

def plot(model, figheight):
    ax = (
        Simulator(model)
        .solve(save_at=np.linspace(0, 5, 101))
        .plot(figsize=(2, figheight))
    )
    ax.legend(frameon=False)
    seaborn.despine(ax.figure)

```

:::: {#fig-code layout-nrow=3 .column-body-outset}

::: {#fig-first-order layout-ncol=3}
```{python}
class Decay(System):
    x: Variable = initial(default=1)
    eq = x.derive() << -x

```

$$
\begin{align}
    \dot{x} &= -x
    \\\\
    x(0) &= 1
\end{align}
$$

```{python}
# | echo: false
plot(Decay, 1)

```

First-order system of an exponential decay.
:::

::: {#fig-second-order layout-ncol=3}
```{python}
class Oscillator(System):
    x: Variable = initial(default=1)
    v: Derivative = x.derive(initial=0)
    eq = v.derive() << -x

```

$$
\begin{align}
    \ddot{x} = -x
    \\\\
    \begin{cases}
        x(0) &= 1 \\
        \dot{x}(0) &= 0
    \end{cases}
\end{align}
$$

```{python}
# | echo: false
plot(Oscillator, 1.5)

```

Second-order system of an harmonic oscillator.
:::

::: {#fig-composition layout-ncol=3}
```{python}
class BigModel(System):
    x: Variable = initial(default=1)
    subsystem = Decay(x=x)
    independent = Decay(x=2)

```

$$
\begin{align}
    \begin{cases}
        \dot{x} = -x \\
        \dot{y} = -y
    \end{cases}
    \\\\
    \begin{cases}
        x(0) &= 1 \\
        y(0) &= 2
    \end{cases}
\end{align}
$$

```{python}
# | echo: false
plot(BigModel, 1.5)

```

First-order system of two exponential decays.
:::
::::

Poincare is a package to define and simulate dynamical systems.
It provides a `System` class,
where we can define `Constant`s, `Parameter`s, `Variable`s,
and create equations linking a variable's derivative with an expression (@fig-first-order).
It also allows to define higher-order systems
by assigning an initial condition to a `Derivative` (@fig-second-order).

@@

1. the variable name to which a component is assigned can be automatically saved in the component for introspection (i.e., `Oscilator.x.name == "x"`),
2. it provides a namespace such that we can easily define multiple independent models in the same script,
3. it allows IDEs to provide autocomplete and refactoring capabilities (`Oscillator.<TAB>` shows `x`, `v` and `eq`),
4. it allows to create instances which can be composed into a bigger model (@fig-composition).

@@
